---
title: "Tour de trackeR"
author: "[Hannah Frick](http://www.ucl.ac.uk/~ucakhfr) and [Ioannis Kosmidis](http://www.ucl.ac.uk/~ucakiko/)"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tour de trackeR}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---


The **trackeR** package provides infrastructure for handling running and cycling
data from GPS-enabled tracking devices. A short demonstration of its functionality is
provided below, based on data from running activities. A more comprehensive introduction
to the package can be found in the vignette "Infrastructure for Running and
Cycling Data", which can be accessed by typing

```{r, eval = FALSE}
vignette("trackeR", package = "trackeR")
```


## Reading data

**trackeR** can currently import files in the Training Centre XML (TCX) format and
.db3 files (SQLite databases, used, for example, by devices from GPSports) through the corresponding
functions *readTCX*() and *readDB3*().

```{r, message = FALSE}
library("trackeR")
filepath <- system.file("extdata", "2013-06-08-090442.TCX", package = "trackeR")
runDF <- readTCX(file = filepath, timezone = "GMT")
```

These read functions return a `data.frame` of the following structure

```{r}
str(runDF)
```

That `data.frame` can be used as an input to the constructor function for **trackeR**'s *trackeRdata* class, to produce a session-based and unit-aware object that can be used for further analyses.

```{r}
runTr0 <- trackeRdata(runDF)
```

The *readContainer*() function combines the two steps of importing the data and constructing the *trackeRdata*
object.

```{r}
runTr1 <- readContainer(filepath, type = "tcx", timezone = "GMT")
identical(runTr0, runTr1)
```

The *readDirectory*() function can be used to read `all` supported files in a directory and produce the corresponding *trackeRdata* objects.


## Visualisations

The package includes an example data set which can be accessed through

```{r}
data("runs", package = "trackeR")
```

The default behaviour of the *plot* method for *trackeRdata* objects
is to show how heart rate and pace evolve over the session.

```{r, fig.width = 7.2, fig.height = 4}
plot(runs, session = 1:7)
```

The elevation profile of a training session is also accessible, here along with the pace.

```{r, fig.width = 7.2, fig.height = 4}
plot(runs, session = 26, what = c("altitude", "pace"))
```

The route taken during a training session can also be plotted on maps from various sources
e.g., from OpenStreetMap. This can be done either on a static map

```{r, message = FALSE, fig.width = 6, fig.height = 6}
plotRoute(runs, session = 4, zoom = 13, source = "osm")
```

or on an interactive map.

```{r, fig.width = 7, fig.height = 5}
leafletRoute(runs, session = 8:15)
```

## Session summaries

The summary of sessions includes basic statistics like duration,
time spent moving, average speed, pace, and heart rate. The speed threshold used to
distinguish moving from resting can be set by the argument *movingThreshold*.

```{r}
summary(runs, session = 1, movingThreshold = 1)
```

It is usually desirable to visualise summaries from multiple sessions. This can be done using the *plot* method for summary objects. Below, we produce such a plot for average heart rate, average speed, distance, and duration.

```{r, fig.width = 7, fig.height = 6.3}
runsSummary <- summary(runs)
plot(runsSummary, group = c("total", "moving"),
  what = c("avgSpeed", "distance", "duration", "avgHeartRate"))
```

The timeline plot is useful to visualise the date and time that the sessions took place and provide information of their relative duration
```{r, fig.width = 7, fig.height = 6.3}
timeline(runsSummary)
```

## Time in zones

The time spent training in certain zones, e.g., speed zones, can also be calculated and visualised.

```{r, fig.width = 7, fig.height = 4}
runZones <- zones(runs[1:4], what = "speed", breaks = c(0, 2:6, 12.5))
plot(runZones)
```

## Quantifying work capacity via W' (W prime)

*trackeR* can also be used to calculate and visualise the work capacity W' (pronounced as `W prime`).
The comprehensive vignette "Infrastructure for Running and Cycling Data" provides the definition of work capacity and details on the *version* and *quantity* arguments.

```{r, fig.width = 7, fig.height = 4}
wexp <- Wprime(runs, session = 11, quantity = "expended", cp = 4, version = "2012")
plot(wexp, scaled = TRUE)
```
