---
title: "Tour de trackeR"
author: "[Hannah Frick](http://www.ucl.ac.uk/~ucakhfr) and [Ioannis Kosmidis](http://www.ucl.ac.uk/~ucakiko/)"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tour de trackeR}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---


The **trackeR** package provides infrastructure for handling running and cycling
data from GPS-enabled tracking devices. A short tour of its functionality is
given below through an analysis of running data. A more comprehensive introduction
to the package can be found in the vignette "Infrastructure for Running and 
Cycling Data" accessible via

```{r, eval = FALSE}
vignette("trackeR", package = "trackeR")
```


## Reading data

**trackeR** currently supports files in the Training Centre XML (TCX) format and
.db3 files (for SQLite, used, e.g., by devices from GPSports) through the corresponding
functions *readTCX*() and *readDB3*().

```{r, message = FALSE}
library("trackeR") 
filepath <- system.file("extdata", "2013-06-08-090442.TCX", package = "trackeR")
runDF <- readTCX(file = filepath, timezone = "GMT")
```

These reading function return a `data.frame` of the following structure

```{r}
str(runDF)
```

which can be fed to the constructor function for **trackeR**'s session-based and
unit-aware data class, *trackeRdata*, for tracking data.

```{r}
runTr0 <- trackeRdata(runDF)
```

These two steps can be combined in the *readContainer*() function for reading GPS
container files of a supported type. All files of a supported type can be read and
turned into *trackeRdata* objects via *readDirectory*().

```{r}
runTr1 <- readContainer(filepath, type = "tcx", timezone = "GMT")
identical(runTr0, runTr1)
```




## Visualisations

The package includes an example dataset which can be accessed through

```{r}
data("runs", package = "trackeR")
```

The *plot* method for *trackeRdata* objects shows how heart rate and pace
(the defaults) evolve over time. 

```{r, fig.width = 7.2, fig.height = 4}
plot(runs, session = 1:7)
```

The elevation profile of a training session is also accessible, here along with the pace.

```{r, fig.width = 7.2, fig.height = 4}
plot(runs, session = 26, what = c("altitude", "pace"))
```

The route taken during a training session can be plotted with different background maps,
e.g., from OpenStreetMap. This can be done either on a static map

```{r, message = FALSE, fig.width = 6, fig.height = 6}
plotRoute(runs, session = 4, zoom = 13, source = "osm")
```

or on an interactive map.

```{r, fig.width = 7, fig.height = 5}
leafletRoute(runs, session = 22:23)
```

## Session summaries

The summary of sessions includes basic statistics like duration,
time spent moving, average speed, pace, and heart rate. The speed threshold used to
distinguish moving from resting can be set by the argument *movingThreshold*.

```{r}
summary(runs, session = 1, movingThreshold = 1)
```

For summaries of multiple session, a plot is usually helpful, here for average
heart rate, average speed, distance, and duration.

```{r, fig.width = 7, fig.height = 6.3}
runsSummary <- summary(runs)
plot(runsSummary, group = c("total", "moving"),
  what = c("avgSpeed", "distance", "duration", "avgHeartRate"))
```

## Time in zones

The time spent training in certain zones, e.g., speed zones, can also be calulated and visualised.

```{r, fig.width = 7, fig.height = 4}
runZones <- zones(runs[1:4], what = "speed", breaks = c(0, 2:6, 12.5))
plot(runZones)
```

## Quantifying work capacity via W' (W prime)

The work capacity quatified as W' (spoken W prime) can also be calculated and visualised.
See the comprehensive vignette "Infrastructure for Running and Cycling Data" for more
details on the *version* and *quantity* arguments. 

```{r, fig.width = 7, fig.height = 4}
wexp <- Wprime(runs, session = 11, quantity = "expended", cp = 4, version = "2012")
plot(wexp, scaled = TRUE)
```
